---
title: "Optimization"
author: "Amey Joshi"
date: "11/23/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Contents

- What is optimization?
- Convex sets
- Convex programming
- Linear programming
- Quadratic programming

## What is optimization?

- To optimize means to make most of what is available.
- One likes to 
  - **maximize** returns,
  - **minimize** risks,
  - **minimize** cost,
  - **maximize** profit,
  - find the **best** route from here to there.
- Sometimes we have to achieve some of these simultaneously,
  - **Maximize** returns but **minimize** risk.
- Many times one has to work under constraints.
  - **Maximize** returns but don't invest more than $5\%$ of my funds in a single instrument,
  - **Maximize** profits but operate within a given inventory of raw materials.

## A general optimization problem

- minimize $f(x)$ subject to
  - $g_i(x) \ge 0$, $i = 1, \ldots, m$,
  - $h_i(x) = 0$, $i = 1, \ldots, n$.
- $f, g_i, h_i$ are arbitrary functions.
- $f$ is called the **objective function** while $g_i$ and $h_i$ are called
**constraints**.
- $x$ is an $n$-dimensional vector. That is, $x = (x_1, \ldots, x_n)$.
- This problem is usually quite difficult. 
- It has been attempted since at least $300$ years and has ushered many new concepts
in Mathematics.
- (Unfortunately) We are not mathematicians.
  - We will talk in the mathematical language but we will focus on the ideas, not
  the technicalities.
  - We will consider the optimization problem in simpler settings.

## Some notation

- Mathematics is a language. Here's some vocabulary.
- $\mathbb{N}$ denotes the set of all natural numbers $0, 1, \ldots$.
- $\mathbb{Z}$ denotes the set of all integers $\ldots, -2, -1, 0, 1, 2, \ldots$.
- $\mathbb{R}$ denotes the set of all real numbers.
- $\mathbb{R}^n$ is a set of all $n$-dimensional vectors whose components are real
numbers. They contain elements like $x = (x_1, \ldots, x_n)$.
- If $S$ is a set and it contains an element $a$, we write it as $a \in S$.

## Convex sets

- Suppose $x, y \in S$ then the set of points $\lambda x + (1 - \lambda)y$, where
$0 \le \lambda \le 1$ is a real number is called the set of all *convex combinations*
of $x$ and $y$.
  - In plain language, this set has all the points on the straight line joining $x$ and $y$.
- A set $S$ is called a convex set if convex combinations of all pairs of its points
lies in the set.
  - In plain language, if you take any two points in the set and draw a straight 
  line joining them then the line should lie entirely in the set.
- Examples of convex sets:
  - A circle, a square, an ellipse, a sphere, a box.
- Examples of sets that are not convex.
  - A circle with a hole, heart-shaped set, a torus (medu-vade)

## Convex functions - 1

- A function $f$ is a rule that maps **every** member of a set $D$ to a **unique**
member of a set $R$. It is written as $f: D \mapsto R$. The set $D$ is called the
domain of the function and the set $R$ the range.
- Examples:
  - A rule that maps an integer to its square. Its domain is $\mathbb{Z}$ and its
  range is $\mathbb{N}$.
- Let $f:\mathbb{R}^m \mapsto \mathbb{R}^n$. $f$ is convex if 
$||f(\lambda x + (1 - \lambda y))|| \le \lambda ||f(x)|| + (1 - \lambda) ||f(y)||$.
- What is $||f(x)||$? $f(x) \in \mathbb{R}^n$. It is a vector. $||f(x)||$ is the
magnitude of the vector.
  - The magnitude of a vector $y = (y_1, \ldots, y_n)$ is $||y|| = \sqrt{y_1^2 + \cdots + y_n^2}$.
- If $f$ is convex, $-f$ is concave.
- A function may be convex in some region, concave in others.
- Choose $\lambda = 1/2$. If $\bar{x} = (x + y)/2$, $\overline{f(x)} = (f(x) + f(y))/2$ then
$||f(\bar{x})|| \le ||\overline{f(x)}||$.
  - Function of average is at most the average of the function.
  
## Convex functions - 2
```{r fig.width=12}
par(mfrow = c(1, 2))
x = seq(from = 0.1, to = 5, length.out = 100)

plot(x, (x-2)^2, xlab = 'x', ylab = expression((x-2)^2), main = 'Convex function', type = 'l')
plot(x, log(x), xlab = 'x', ylab = 'log(x)', main = 'Concave function', type = 'l')
```

- In a convex function, a chord lies above the tangent. In a concave function it is it the other way round.

## Convex functions - 3
- Why are we so concerned about convex functions?
  - For a convex function, a local minimum is the global minimum.
  - If you find a minimum you can be assured that it is **the** minimum. You don't have to search any further.
- An optimization problem consists of
  - A set $F$ of feasible points.
  - A cost function $c: F \mapsto \mathbb{R}$.
- The solution of the optimization problem consists in finding $x^\ast \in F$ such that
$c(x^\ast) \le c(x)$ for all $x \in F$.
- A convex optimization problem is the one in which the set $F$ and the function $c$
are convex.
- In general, convex problems are much easier to solve than non-convex ones.
- CPLEX solves convex problems.

## Linear Programming
- Programming?
  - This is a term from the 1940s when a computer meant a person who computes and
  a program mean a plan.
  - In particular, it referred to the US army's training and logistic plans. George
  Dantzing, the man who developed the simplex algorithm was working on military
  problems during WW-II.
- It has the same meaning in 'dynamic programming', a style of building algorithms
that saves intermediate results in tables.
- In a linear programming problem the objective function and the constraints are
**linear functions**.
  - Constraints are in the form of lines/planes/hyperplanes.
  - Feasible region is a convex polygon/polyhedron/polytope.
- An optimum, if it exists, is at the corner of the polytope or along an edge.

## Linear Programming - Enough of theory
- The 'hello world' of optimization is 'the diet problem', a simple version of which is:
```{r}
library(knitr)
diet.data <- data.frame(food = c('Trout', 'CB Sandwich', 'Burrito', 'Hamburger'),
                        Vit.A = c(203, 90, 270, 500),
                        Vit.C = c(92, 84, 80, 90),
                        Vit.D = c(100, 230, 512, 210),
                        calories= c(600, 350, 250, 500))
needs <- data.frame(nutrition = c('Vit.A', 'Vit.C', 'Vit.D'), amount = c(2000, 300, 430))
kables(list(kable(diet.data, caption = 'Foods'),
kable(needs, caption = 'Nutritional needs')))

```

- Minimize your calorie intake but ensure that you get minimum nutrition of 
every kind.
- How to solve it? Let $x_1, x_2, x_3, x_4$ be the amounts of foods consumed.
  - The total calories consumed is $600x_1 + 350x_2 + 250x_3 + 500x_4$,
  - Vit.A consumed is $203x_1 + 90x_2 + 270x_3 + 500x_4$,
  - Vit.C consumed is $92x_1 + 84x_2 + 80x_3 + 90x_4$,
  - Vit.D consumed is $600x_1 + 350x_2 + 250x_3 + 500x_4$.
- Natural to assume $x_1 \ge 0, x_2 \ge 0, x_3 \ge 0, x_4 \ge 0$.

## The diet problem - 1
```{r echo=TRUE}
library(lpSolve)
tot.calories <- c(600, 350, 250, 500)
nutrition <- matrix(data = c(203, 90, 270, 500, 92, 84, 80, 90, 100, 230, 512, 210), byrow = TRUE, nrow = 3)
needs <- c(2000, 300, 430)
food <- c('Trout', 'CB Sandwich', 'Burrito', 'Hamburger')
stopifnot(length(needs) == nrow(nutrition))
lp.results <-
  lp(direction = "min",
     objective.in = tot.calories,
     const.mat = nutrition,
     const.dir = rep(">=", nrow(nutrition)),
     const.rhs = needs)
if (lp.results$status == 0) {
  print(paste('Calories consumed =', round(lp.results$objval, 0)))
  X <- lp.results$solution
  print(paste('Solution:', food[1], ":", X[1], ',', food[2], ':', X[2], ',', food[3], ':', X[3], ',', food[4], ':', X[4]))
}
```

Even professional mathematicians don't order $7.407$ Burritos.

## The diet problem - 2
```{r echo=TRUE}
lp.results <-
  lp(direction = "min",
     objective.in = tot.calories,
     const.mat = nutrition,
     const.dir = rep(">=", nrow(nutrition)),
     const.rhs = needs,
     int.vec = c(1, 2, 3, 4)) # Insist that the solution must be an integer.
if (lp.results$status == 0) {
  print(paste('Calories consumed =', round(lp.results$objval, 0)))
  X <- lp.results$solution
  print(paste('Solution:', food[1], ":", X[1], ',', food[2], ':', X[2], ',', food[3], ':', X[3], ',', food[4], ':', X[4]))
}
```
- Adding the last argument is to the function `lp` is a trivial change to the code but
a significant change to the problem.
  - Integer linear programming is NP-hard.

## Quadratic programming